Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2025-02-17 src/tex/ms.tex
[showyourwork] overleaf sync

\subsection{Neural network emulator for \texttt{CHEMPY}}
Despite the simplifications made by the GCE model \texttt{CHEMPY}, the run-time of \texttt{CHEMPY} and the high-dimensionality of the parameter space incurs some difficulties when sampling the distribution of the global parameters $\vec\Lambda = \{\alpha_\mathrm{IMF},\log_{10}(N_\mathrm{Ia})\}$. 
To alleviate this, we follow \cite{Philcox_2019} and implement a \textit{neural network} (NN) emulator of the \texttt{CHEMPY} simulator. We design the NN as a simple feed-forward neural network with 2 hidden layers and 100 neurons in the first and 40 neurons in the second layer. Training this tiny emulator takes about 50s on the CPU.

@@

Despite the additional complexity introduced by using multiple stellar data-points, our NN simply needs to predict the birth-time abundances for a single star (with index $i$) given a set of six parameters; $\{\vec\Lambda,\vec\Theta_i,T_i\}$. The same NN can be used for all $n_\mathrm{stars}$ stars (and run in parallel), reducing a set of $n_\mathrm{stars}$ runs of \texttt{CHEMPY} to a single matrix computation. With the above network parameter choices, the NN predicts abundances with an absolute percentage error of \variable{output/ape_NN.txt}, which is far below typical observational errors and even smaller away from the extremes of parameter space (see Fig.~\ref{fig:ape_NN}).
Our NN emulator is publicly available on the github repository accompanying this manuscript \href{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train_torch_chempy.py}{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train\_torch\_chempy.py} with pre-trained NN weights available on zenodo \href{https://zenodo.org/records/14507134}{https://zenodo.org/records/14507134}.

@@

% %The larger volume of data should be able to give tighter statistical constraints on those parameters that are held fixed across the galaxy, but complexity is added since we must allow each star to carry its own set of local ISM parameters.

@@

We use a Neural Posterior Estimator (NPE) \cite{zeghal2022neuralposteriorestimationdifferentiable} which utilises the gradients of the generative model $\mathcal{M}$ with a Masked Autoregressive Flow model (MAF) \cite{papamakarios2018maskedautoregressiveflowdensity} containing 10 hidden features and 1 transformation layer for the normalizing flow.

>>>>

\subsection{Neural network emulator for \texttt{CHEMPY}}
Despite the simplifications made by the GCE model \texttt{CHEMPY}, the run-time of \texttt{CHEMPY} and the high-dimensionality of the parameter space incurs some difficulties when sampling the distribution of the global parameters $\vec\Lambda = \{\alpha_\mathrm{IMF},\log_{10}(N_\mathrm{Ia})\}$. 
To alleviate this, we follow \cite{Philcox_2019} and implement a \textit{neural network} (NN) emulator of the \texttt{CHEMPY} simulator. We design the NN as a simple feed-forward neural network with 2 hidden layers and 100 neurons in the first and 40 neurons in the second layer. The NN is trained on $\sim100,000$ data points and validated on $\sim50,000$ additional data points created with \texttt{CHEMPY} using a uniform prior over the $5\sigma$-range of the original Gaussian prior stated in table~\ref{tab:priors}. The batch size is set to 64 and the learning rate is set to 0.001. We train for 20 epochs using a schedule free optimizer \citep{schedulefree}. 
Training this tiny emulator takes about 200s on the CPU.

@@

Despite the additional complexity introduced by using multiple stellar data-points, our NN simply needs to predict the birth-time abundances for a single star (with index $i$) given a set of six parameters; $\{\vec\Lambda,\vec\Theta_i,T_i\}$. The same NN can be used for all $n_\mathrm{stars}$ stars (and run in parallel), reducing a set of $n_\mathrm{stars}$ runs of \texttt{CHEMPY} to a single matrix computation. With the above network parameter choices, the NN predicts abundances with an absolute percentage error of \variable{output/ape_NN.txt}, which is far below typical observational errors and even smaller away from the extremes of parameter space (see Fig.~\ref{fig:ape_NN}). In fact, we will add additional observational uncertainty to our mock observational data later during training of the neural posteriro estimater network.

Our NN emulator is publicly available on the github repository accompanying this manuscript \href{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train_torch_chempy.py}{https://github.com/TobiBu/sbi-chempy/blob/main/src/scripts/train\_torch\_chempy.py} with pre-trained NN weights available on zenodo \href{https://zenodo.org/records/14507134}{https://zenodo.org/records/14507134}.

@@

@@

We use a Neural Posterior Estimator (NPE) \cite{zeghal2022neuralposteriorestimationdifferentiable} which utilises the gradients of the generative model $\mathcal{M}$ with a Neural Spline Flow (NSF) \citet{durkan2019neuralsplineflows} containing 20 hidden features and 10 transformation layers for the normalizing flow. The NSF uses as invertible transformation smooth and flexible spline functions to allow efficient sampling. The expressivity of the spline allows the NDE to be capture complex distributions, while also maintaining computational tractability. The final model was selected after extensive hyperparameter tuning, varying: the architecture between NSF, Masked Autoregressive Flow (MAF) \cite{papamakarios2018maskedautoregressiveflowdensity}, and Mask Autoregressive 
Flow with rational-quadratic spline (MAF-RQS), the number of neurons between {10, 20, 50, 100}, and the number of transformations between {1, 5, 10, 30}, optimizing over the test set for both the highest mean log posterior probability and the best calibration as measured by the TARP value.
